---
title: "Guide to autocorrelated home range estimation"
description: |
    A step-by-step R tutorial supplementing our manuscript "***Autocorrelation-informed home range estimation: a review and practical guide"***" using the 'ctmm' package
author: 
  - name: "Inês Silva"
    orcid_id: 0000-0003-4850-6193
  - name: "Christen H. Fleming"
  - name: "Michael J. Noonan"
  - name: "Jesse Alston"
  - name: "Cody Folta"
  - name: "William F. Fagan"
  - name: "Justin M.Calabrese"
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
        highlight: kate
        code_download: true
        toc: true
        toc_depth: 2
        toc_float: true
bibliography: references/R_tutorial.bib
csl: references/apa.csl
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, 
  fig.retina = 1, fig.width = 9, fig.height = 8
)

```

# Introduction

Home range estimation is a key output from tracking datasets, but the inherent properties of animal movement can lead traditional methods to under- or overestimated their size. **Autocorrelated Kernel Density Estimation (AKDE)** methods were designed to be statistically efficient while explicitly dealing with the complexities and biases of modern movement data, such as *autocorrelation*, *small sample sizes*, and *missing or irregularly sampled data*.

<aside>
*Home range*: the area repeatedly used throughout an animal’s lifetime for all its normal behaviors and activities, excluding occasional exploratory moves.
</aside>

This tutorial is a companion piece to our manuscript *"Autocorrelation-informed home range estimation: a review and practical guide"*. For any definitions, check the main manuscript or the *[Glossary]*. The **AKDE** family of home range estimators will be run using **R software** (<https://www.r-project.org/>) and the `ctmm` package [@calabrese_ctmm_2016]. The techniques and mitigation measures available within this package include:

```{r, include = FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
library(ctmm)

library(here)

# if necessary, quickly load already run objects
load(here::here("files/AKDE_R-tutorial_ctmm-runs.RData"))

table.dat <- read.csv(here::here("files/AKDE_R-tutorial_summary-table.csv"))

colnames(table.dat) <- c("Method",
                         "When to run?", 
                         "What does it do?", 
                         "R function")

table <- kable(table.dat, booktabs = TRUE) %>% 
 row_spec(0, bold = TRUE) %>% 
 column_spec(4, monospace = T) %>% 
 kable_styling(latex_options = c("striped", "hold_position"))

```

```{r, echo = FALSE}

table

```

AKDE~c~ and pHREML are default arguments within the `akde()` and `ctmm.select()` functions, respectively: both will run automatically if arguments `debias` and `method` are left unspecified. For most situations, we recommend keeping both of these arguments as the default.


```{r}

# Installing & loading package:
install.packages("ctmm")
library(ctmm)

```

------------------------------------------------------------------------

# Data Preparation

We will use two datasets, both available within the `ctmm` package: African buffalos (*Syncerus caffer*), and Mongolian gazelles (*Procapra gutturosa*). Information on the data collection protocol is available in Cross *et al.* [-@cross_disease_2009] and Fleming *et al.* [-@fleming_fine-scale_2014]. The `ctmm` package requires data to conform to Movebank naming conventions (<https://www.movebank.org/node/2381>). We recommend uploading your data to Movebank (<http://www.movebank.org/>) as this will facilitate data preparation, and ensure that your data are correctly formatted for `ctmm`. If needed, Movebank allows you to keep your data private.

We will focus on the simplest data structure:

- `animal ID` --- An individual identifier for each animal tracked;
- `timestamp` or `t` --- The date and time corresponding to a sensor measurement;
   - **Example**: 2021-01-01 18:31:00.000
      - **Format**: yyyy-MM-dd HH:mm:ss.SSS
- `longitude` or `x` --- The geographic longitude of the location as estimated by the sensor.
   - **Example**: -121.1761111
      - **Units**: decimal degrees, WGS84 reference system.
- `latitude` or `y` --- The geographic latitude of the location as estimated by the sensor;
   - **Example**: -41.0982423
      - **Units**: decimal degrees, WGS84 reference system.

Location can also be described as UTM locations instead of latitude/longitude. In this case, you should provide `UTM easting`, `UTM northing`, and `UTM zone`. For all terms and conventions, please see the full vocabulary list here: <http://vocab.nerc.ac.uk/collection/MVB/current/>.

## **Step 1.** -- Formatting and loading an animal tracking dataset

You can import data into R through the `read.table()` or `read.csv()` functions; make sure to navigate to the appropriate folder or working directory. For this tutorial, our data is already prepared into a list of `telemetry` objects which we can load into R:

### 1.1. Buffalo tracking data

```{r}

data("buffalo")
animal1_buffalo <- buffalo[[4]] # select individual number 4
head(animal1_buffalo)

# Plotting locations:
plot(animal1_buffalo)

```

This dataset showcases an irregular sampling schedule: the buffalo nicknamed *"Pepper"* had a sampling rate shift from one fix every hour to one fix every two hours. We will use this dataset to highlight data irregularity and the **wAKDE** mitigation measure.

### 1.2. Gazelle tracking data

```{r}

data("gazelle")
animal2_gazelle <- gazelle[[11]] # selecting individual number 11
head(animal2_gazelle)

# Plotting locations:
plot(animal2_gazelle)

```

<aside>
*Home range crossing time*: the time required for an animal to cross the linear extent of its home range.
</aside>

Mongolian gazelles have a ***home range crossing time*** of a few months, and with a maximum longevity around 10 years, it is impossible to get a considerable ***effective sample size*** no matter the study duration [@fleming_overcoming_2019]. We will use this dataset to highlight how to check ***effective sample size*** and apply the **parametric bootstrap** mitigation. 

<aside>
*Effective sample size (N)*: number of range crossings that occurred during the observation period. Can be roughly estimated by dividing the duration of the tracking dataset by the average home range crossing time parameter.
</aside>

------------------------------------------------------------------------

# *Data Analysis*

## **Step 2.** -- Checking for the *range residency* assumption

First, we want to check if our first tracking dataset (`animal1_buffalo`) can be used for home range estimation by checking for ***range residency***. To achieve this, we calculate the **semi-variance function (SVF)**, and visualize it through the `variogram()` function.

<aside>
*Range residency*: the tendency of an animal to remain within their home range.
</aside>

**Variograms** are an unbiased way to visualize *autocorrelation* structure, representing the average square displacement (y-axis) over a specific time lag (x-axis). To facilitate interpretation, we have the **SVF** of `animal1_buffalo` zoomed out (right) to showcase all time lags and (left) zoomed in to showcase time lags up to two months:

```{r}

level <- 0.95 # we want to display 95% confidence intervals
xlim <- c(0,2 %#% "month") # to create a window of 2 months

SVF <- variogram(animal1_buffalo)
par(mfrow = c(1,2))
plot(SVF, fraction = 1, level = level)
abline(v = 1, col = "red", lty = 2) # adding a line at 1 month 
plot(SVF, xlim = xlim, level = level)
abline(v = 1, col = "red", lty = 2)

```

We can see that the variogram flattens (*i.e.*, reaches an asymptote) after approximately **1 month** (<span style="color: red;">red</span> line). This also indicates at how coarse the timeseries needs to be to assume independence (no autocorrelation), and corresponds to when traditional methods ---such as **minimum convex polygons (MCPs)** and **Kernel Density Estimators (KDEs)**--- could be applied without violating their assumptions.

## **Step 3.** -- Selecting the best-fit movement model through *model selection*

It is necessary to choose a home range estimator that accounts for the autocorrelated structure of the data, now that we see that it is **not** independently and identically distributed (non-IID). We need to test what movement model may explain the autocorrelated structure of our tracking data. We can run different movement processes with **maximum likelihood (ML)** or other parameter estimators, such as **perturbative Hybrid REML (pHREML)**. To facilitate further comparisons, we will run both ML and pHREML with the `ctmm.select` function. 

```{r}

# Calculate an automated model guesstimate:
GUESS1 <- ctmm.guess(animal1_buffalo, interactive = FALSE)

# Automated model selection, starting from GUESS:
FIT1_ML <- ctmm.select(animal1_buffalo, GUESS1, method = 'ML')
FIT1_pHREML <- ctmm.select(animal1_buffalo, GUESS1, method = 'pHREML')
## reminder: it will default to pHREML if no method is specified.

summary(FIT1_ML)
summary(FIT1_pHREML)

```

Within these summaries, `$name` provides the selected best-fit model, `$DOF` provides information on the degrees of freedom (where `$DOF["area"]` corresponds to the ***effective sample size*** of the home-range area estimate), and `$CI` are the parameter outputs (area, position autocorrelation timescale, velocity autocorrelation timescale, and speed).

The typical pool of candidate models includes isotropic (when diffusion is the same in every direction; symmetrical) and anisotropic (when diffusion varies with direction; asymmetrical) variants. The automated model selection shows that *OUF anisotropic* (anisotropic Ornstein-Uhlenbeck foraging process) is our best-fit model. This movement process features a home range, correlated positions, and correlated velocities. To check the full model selection table, we can run the following command:

```{r}

FIT1_pHREML_verbose <- ctmm.select(animal1_buffalo, GUESS1, verbose = TRUE)
summary(FIT1_pHREML_verbose)

```

By adding the argument `verbose = TRUE` we have access to the model selection table. By default, model selection is based on *Akaike’s Information Criterion adjusted for small sample sizes* (AICc). The `ctmm` package also offers BIC, LOOCV, and HSCV. LOOCV seems to work slightly better for very small datasets, but we recommend AICc for the majority of datasets.

## **Step 4.** -- Feeding a movement model into the *home range estimator*

Now we can fit this movement process into the `akde()` function, and estimate the home range of `animal1_buffalo`. This function currently defaults to the **area-corrected AKDE**, or **AKDEc** (Fleming & Calabrese 2017):

```{r}

# Run an area-corrected AKDE:
UD1_ML <- akde(animal1_buffalo, FIT1_ML)
UD1_pHREML <- akde(animal1_buffalo, FIT1_pHREML)

summary(UD1_pHREML)$CI # home range area estimation

```

We have calculated our home range for `animal1_buffalo`, resulting in an estimation of 757 km^2^ (with 95% confidence intervals: 430–1,175 km^2^).

## **Step 5.** -- Evaluating additional *biases*, applying *mitigation measures*

### 5.1. Buffalo tracking data

```{r}

summary(UD1_pHREML)$DOF["area"] # effective sample size of animal1
nrow(animal1_buffalo) # absolute sample size

```

Our output here also reveals more information regarding our dataset: the ***effective sample size (N)*** and the ***absolute sample size (*n*)***. We can return this measure with the `summary` function: in our case, the *N* for `animal1_buffalo` is 15.7. Comparatively, our ***absolute sample size*** is easy to output, as it is the total number of observations within our dataset (*n* = 1,725).

<aside>
*Absolute sample size (n)*: the number of observations (fixes) in a dataset.
</aside>

As mentioned earlier, `animal1_buffalo` had a device malfunction that led GPS fixes to shift from one fix per hour, to one fix every two hours. As such, this individual is particularly suited for a **weighted AKDEc** (or **wAKDEc**), so we can re-run the function with weights set to `TRUE`:

```{r}

UD1w_pHREML <- akde(animal1_buffalo, FIT1_pHREML, weights = TRUE)
summary(UD1w_pHREML)$CI # home range area estimation (weighted)

```

Our new home range area estimation for `animal1_buffalo` is 761 km^2^ (with 95% confidence intervals: 432–1,182 km^2^). We can now plot our home range estimate for `animal1_buffalo`:

```{r}

# Creating an extent that includes both UDs at the 95% CI level:
EXT <- extent(list(UD1_ML, UD1_pHREML, UD1w_pHREML), level = 0.95)

# Plotting pHREML (with and without weights) side-by-side:
par(mfrow = c(1,2))
plot(animal1_buffalo, UD = UD1_pHREML, ext = EXT)
title(expression("pHREML AKDE"["C"]))
plot(animal1_buffalo, UD = UD1w_pHREML, ext = EXT)
title(expression("pHREML wAKDE"["C"]))

```

For `animal1_buffalo`, the difference between model parameter estimators is not substantial; we only have a ~5.7% AKDE area underestimation by ML compared to pHREML. However, the data fits the spatial locations much better. 

```{r}

( 1 - summary(UD1_ML)$CI[1,2] / summary(UD1w_pHREML)$CI[1,2] ) * 100

```

### 5.2. Gazelle tracking data

We can also check the difference with `animal2_gazelle`’s tracking data, where the small ***effective sample size*** issue is clearer: 

```{r}

GUESS2 <- ctmm.guess(animal2_gazelle, interactive = FALSE)

FIT2_ML <- ctmm.select(animal2_gazelle, GUESS2, method = 'ML')
FIT2_pHREML <- ctmm.select(animal2_gazelle, GUESS2, method = 'pHREML')

UD2_ML <- akde(animal2_gazelle, FIT2_ML)
UD2_pHREML <- akde(animal2_gazelle, FIT2_pHREML)

```

With `animal2_gazelle`, we have a more substantial area underestimation by ML compared to pHREML (~15.2%). We can also see that our ***effective sample size*** is only 4.5, with an ***absolute sample size*** of 49 (*N* $\ll$ *n*). 

```{r}

( 1 - summary(UD2_ML)$CI[1,2] / summary(UD2_pHREML)$CI[1,2] ) * 100

summary(UD2_pHREML)$DOF["area"] # effective sample size
nrow(animal2_gazelle) # absolute sample size

```

At this point, we have selected a movement process, fed it into a home range area estimation with different model parameter estimators, and corrected for irregular sampling rates. With small ***effective sample sizes***, it is important to see if **parametric bootstrapping** may be worth it to further reduce our estimation error. In order to do so, we can check the expected order of bias from pHREML:

```{r}

# Expected order of pHREML bias:
1/summary(FIT2_pHREML)$DOF['area']^2

```

The bias is currently $\mathcal{O}(5\%)$ ("in the order of" 5%). As such, we will run parametric bootstrapping for `animal2_gazelle`. The relative error target is 1% by default (`argument error = 0.01`), but can be adjusted if necessary.

```{r}

start_time <- Sys.time() # start recording running time
BOOT <- ctmm.boot(animal2_gazelle, FIT2_pHREML, trace = 2)
## note: this function incurs substantial computational cost, may take hours.
( total_time <- Sys.time() - start_time ) # output running time

summary(BOOT)
1/summary(BOOT)$DOF['area']^3 # expected order of bias

```

We can see that the expected order of bias was reduced to 2.3%, which is comparable to the numerical error target of 1%. To reduce the numerical error further, we would need to change the default relative error target of `ctmm.boot`, but the computational cost would continue to increase, and the comparably large statistical bias (2%) would remain.

Now we will calculate the **AKDEc** based on the estimated parameters, and plot the home range of `animal2_gazelle`. Because of small ***effective sample size***, we set optimal weights to `TRUE` for improved statistical efficiency:

```{r}

UD2_bpHREML <- akde(animal2_gazelle, BOOT, weights = TRUE)
summary(UD2_bpHREML)$CI

```

Finally, we have calculated our home range for `animal2_gazelle`, with an estimated area of 13,274 square kilometers (with 95% confidence intervals: 3,231–30,280 km^2^). Our uncertainty with `animal2_gazelle` is substantially higher than with `animal1_buffalo`, as expected due to the small ***effective sample size***.

```{r}

# Creating an extent that includes both UDs at the 95% CI level:
EXT <- extent(list(UD2_pHREML, UD2_bpHREML), level = 0.95)

# Plotting pHREML and bootstrapped-pHREML side-by-side:
par(mfrow = c(1,2))
plot(animal2_gazelle, UD = UD2_pHREML, ext = EXT)
title(expression("pHREML AKDE"["C"]))
plot(animal2_gazelle, UD = UD2_bpHREML, ext = EXT)
title(expression("Bootstrapped pHREML wAKDE"["C"]))

```

The results presented here were generated with `R` version `4.0.5`, and `ctmm` version `0.6.0`.

------------------------------------------------------------------------

# Glossary
<a id="glossary"></a>

-- *Home range*: the area repeatedly used throughout an animal's lifetime for all its normal behaviors and activities, excluding occasional exploratory moves.

-- *Range residency*: the tendency of an animal to remain within their home range.

-- *Home range crossing time*: the time required for an animal to cross the linear extent of its home range.

-- *Absolute sample size* (*n*): the number of observations (fixes) in a dataset.

-- *Effective sample size* (*N*): number of range crossings that occurred during the observation period. Can be roughly estimated by dividing the duration of the tracking dataset by the average *home range crossing time* parameter.
